Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-11 paper.tex
Add some background motivational stuff on ARGs

Computing the likelihood of a putative ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
decomposes the likelihood of an ARG under the coalescent with recombination
into a product of the probablility of 
each common ancestor event, each recombination event and the probablility of
the inter-event waiting times. In essence, computing the KYF likelihood requires
``replaying'' the full state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the probability of the configuration of lineages and distribution
of ancestral material at each time point. It therefore requires complete
detail about each of these events to be stored. 
Such complete detail is not present in many estimated
ARGs~\citep{wong_general_2023} or even in most simulations.
Indeed, supporting likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to the \texttt{msprime} simulator~\citep{baumdicker_efficient_2021}.
More recent approaches to defining likelihoods on such models require
similarly exhaustive input\citep[e.g.][]{guo_recombination-aware_2022} (MORE?
Anything for GC in bacteria, etc?)

>>>>

Computing the likelihood of a putative ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
In essence, computing the KYF likelihood requires
``replaying'' the  state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the exact configuration of lineages and distribution
of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
It therefore requires complete
detail about each of these events to be stored. 
Other approaches to defining the likelihood of an ARG under a coalescent-like 
model follow the same basic event-by-event 
strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
[Are there others?  The bacterial stuff from Didelot etc seem to focus on
a tree-by-tree likelihood, rather than global graph, so not clear it's the 
same thing]
Computing the likelihood under the SMC rather than the Hudson coalescent 
simplifies bookkeeping somewhat, 
but a complete event-by-event description of the 
input ARG is still required.

Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which contain no information about recombination events (other than their
breakpoints) and where internal nodes marking coalesces are unlabelled.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
contain the level of detail required to compute a likelihood under
the KYF approach.
